---
title: An R Markdown document converted from "Meg_Applying_Fun_R.ipynb"
output: html_document
---

# Applying functions in R

Installing to tead Excel files:

```{r}
# install.packages('rio') #installation was not required as it was done in Assignment 1
```

# reading data

```{r}
# get the link for where the excel file is located
linkGit="https://github.com/megvonguru/HW-2/raw/refs/heads/main/FSI-2023-DOWNLOAD.xlsx"


# library(rio) # package needed
# from the link above, import the file and save it under the name fragility23
fragility23=rio::import(file = linkGit) #object that will hold the result
```

```{r}
# display the content of fragility 23 in a readable format 
str(fragility23)
```

## Apply square root function?

```{r}
# whole DF?
# sqrt(fragility23)
```

```{r}
# several columns (only numerical)

# calculate the square root for columns 4 and 5
sqrt(fragility23[,4:5])
```

```{r}
# one column
# calculate the square root of each value in the column total
sqrt(fragility23$Total)
```

```{r}
# one value
# calculate the square root of the first element of the column total
sqrt(fragility23$Total[1])
```

## Applying **sum()**:

```{r}
# sum everything
# calculate the sum of all values for columns 4 and 5
sum(fragility23[,4:5])
```

```{r}
# sum by column
# calculate the sum of the columns 4 and 5 and then print the data
print(apply(fragility23[,4:5],2,sum))
```

```{r}
#notice
# find the sum of columns 4 and 5 and then identify the type of data that is returned 
typeof(apply(fragility23[,4:5],2,sum))
```

If you do not see **list**, then it is a vector. ⏫

```{r}
# sum by row
# calculate the sum of rows 4 and 5 and then print the output
print(apply(fragility23[,4:5],1,sum))
```

### Apply by iterating:

```{r}
# calculate the sum of all values in the 4th and 5th column and then print them as a list
print(lapply(fragility23[,4:5],sum))
```

Notice output of **lapply**:

```{r}
# apply the sum function on columns 4 and 5 and identify the type of internal storage of the returned list
typeof(lapply(fragility23[,4:5],sum))
```

```{r}
# apply the sum function on columns 4 and 5 and identify the type of high level storage of the returned list
class(lapply(fragility23[,4:5],sum))
```

Notice output of **sapply**:

```{r}
# calculate the sum of the elements of the columns in 4 and 5 and print the results in a vector format
print(sapply(fragility23[,4:5],sum))
```

```{r}
# calculate the sum of the elements of the columns in 4 and 5 and identify the high storage type
class(sapply(fragility23[,4:5],sum))
```

Similarly:

```{r}
# do the square root of the elements of the columns in 4 and 5 and then print the list
print(lapply(fragility23[,4:5],sqrt))
```

```{r}
# do the square root of the elements of the columns in 4 and 5 and then return the data type of the list
class(lapply(fragility23[,4:5],sqrt))
```

```{r}
# do the square root of the elements of the columns in 4 and 5 and then print the data in the vector format
print(sapply(fragility23[,4:5],sqrt))
```

```{r}
# do the square root of the elements of the columns in 4 and 5 and then return the type of the vector
class(sapply(fragility23[,4:5],sqrt))
```

Now our own function:

```{r}
# create a function called theOnesOK with input parameter DF_country_and variable
theOnesOK = function(DF_country_and_variable) {
  # take the 2nd column of the input list and save it under variable_values
  variable_values <- DF_country_and_variable[,2]
  # calculate the average of the elements that do not have any missing data and save it under avg_value  
  avg_value <- mean(variable_values, na.rm = TRUE)
  # if the value is greater than the average, label it as Above Average. Otherwise label it as Below/At Average. 
  is_above <- ifelse(variable_values > avg_value, "Above Average", "Below/At Average")
  # create a new column called Status and store all the Above Average data
  DF_country_and_variable$Status <- is_above
  # return the orginal list with the new status column 
  return(DF_country_and_variable)
}
```

```{r}
# call the function theOnesOK with columns Country and Demographic Pressures 
theOnesOK(fragility23[,c('Country','S1: Demographic Pressures')])
```

# the function to compute the row wise average across the given columns and return the new data frame with country column and the average
```{r}
# create a function with the name mystery with input parameters data frame, vector of columns and the country column which has as a default value of country
mystery=function(DF,positionsToUse,CountryColumn='Country'){
  # extract the country column from DF and ignore it if its not a vector
  newDF=DF[,c(CountryColumn),drop = FALSE]
  # create a new column named average 
  average='average'
  # calculate the average for each row and save it under the new column average
  newDF[,average]=apply(DF[,positionsToUse],1,mean,na.rm = TRUE)
  # return the country identifier as well as row wise average
  return(newDF[,c(CountryColumn,average)])
}
```

```{r}
# extract the specific columns total and S1
fragility23[,c('Total',	'S1: Demographic Pressures')]
```
# New Function theOnesOK2
# function will label each row as above or below the average based on the chosen variable
# in the original theOnesOK function, the row was being labeled as above or below based on column 2 but not on the chosen variable
```{r}
# create a function by the name theOnesOk2 
theOnesOK2 = function(DF, DFvariable, CountryColumn='Country') {
  # extract the variable and save it in variable_values
  variable_values <- DF[,DFvariable]
  # calculate the average of the elements listed in variable values
  avg_value <- mean(variable_values, na.rm = TRUE)
  # if the value is greater than the average, label it as Above Average. Otherwise label it as Below/At Average. 
  is_above <- ifelse(variable_values > avg_value, "Above Average", "Below/At Average")
  # create a new column name as Status_on <column name>
  newname = paste('Status_on', DFvariable)
  # add the new status into the data frame 
  DF[,newname] <- is_above 
  # return the country identifier along with the status column 
  return(DF[,c(CountryColumn, newname)])
}
```

